---
title: "Speed and Maintainability Optimizations"
author: "geo maintainers"
date: "2026-01-28"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Speed and Maintainability Optimizations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

This vignette documents the recent performance and maintainability work
focused on `geolines()` and the clipping pipeline used by map drawing. The
changes were driven by profiling results that showed clipping dominated runtime
for large line datasets.

Key improvements:

- A fast rectangle clipper used by `geolines()` in the common case where the
  plot border is axis-aligned.
- A new C++ polygon clipper for non-rectangular borders.
- Reduced allocation and early-out checks in the legacy R clipping path.
- An opt-in cache in `geolines()` for repeated draws with identical inputs.
- Lightweight profiling and benchmark scripts in `tests/perf/`.

## Rectangle clipping fast path

`geolines()` typically clips lines against the plot border, which is an axis-
aligned rectangle derived from `geopar$limx` and `geopar$limy`. The new path
recognizes this case and uses a Liang–Barsky style segment clipper to handle
segments in O(1) time per segment.

- R fallback: `geo_clip_polyline_to_rect()`
- C++ implementation: `geo_clip_polyline_to_rect_cpp()`

The C++ path is used automatically when available.

## Polygon clipping in C++

For non-rectangular borders (e.g., irregular masks), the package now uses a C++
implementation of the existing R algorithm:

- C++ implementation: `geo_clip_polyline_to_polygon_cpp()`
- R wrapper with fallback: `geo_clip_polyline_to_polygon()`

The C++ version mirrors the R logic, including polygon closure and segment
splitting at intersections, while reducing overhead from repeated allocations.

## Allocation reductions and early-outs

The R implementation was adjusted to reduce vector growth and to skip segments
that cannot intersect the polygon based on bounding boxes. These improvements
remain as a fallback when C++ is not available and aid maintainability by
keeping the R version clear and consistent with the C++ logic.

## Optional caching in `geolines()`

`geolines()` now accepts a `cache` flag. When `cache = TRUE`, the function
stores the last `(projected line, border)` pair and reuses the clipped result
if an identical request is repeated.

This is intended for interactive workflows where the same geometry is drawn
multiple times with identical inputs.

## How to reproduce profiling and benchmarks

The repository includes small scripts in `tests/perf/` that exercise common
paths. These are not run by default during testing.

```{r profile-geolines, eval=FALSE}
# Install into a temp library
R
CMD
INSTALL - l / tmp / geo - lib
.

# Profile geolines in a typical rectangular clipping scenario
Rscript
tests / perf / profile - geolines.R / tmp / geo - lib
```

```{r bench-geolines, eval=FALSE}
# Quick timing of geolines with the rectangle clipper
Rscript
tests / perf / bench - geolines.R / tmp / geo - lib
```

```{r bench-polygon, eval=FALSE}
# Benchmark polygon clipping
Rscript
tests / perf / bench - clip - polygon.R / tmp / geo - lib

# Compare C++ vs local R reference implementation
Rscript
tests / perf / bench - clip - polygon - compare.R / tmp / geo - lib
```

## Example local results

The following numbers are from local runs on a single developer machine and are
included to show the magnitude of improvement, not as guaranteed performance.

| Scenario | Metric | Before (R rect clip) | After (C++ rect clip) |
|---|---|---:|---:|
| `geolines()` benchmark | Median elapsed (n=10) | 0.3265s | 0.0470s |
| `geolines()` profile | Total time (10 draws) | 3.33s | 0.48s |

| Scenario | Metric | R polygon clip | C++ polygon clip |
|---|---|---:|---:|
| Polygon clip benchmark | Median elapsed (n=25) | 24.0 ms | <0.5 ms (<500 us) |
| Polygon clip benchmark | Mean elapsed (n=25) | 26.0 ms | 0.12 ms (120 us) |
| Polygon clip benchmark | Mean speedup | 1x | ~217x |

## Notes on correctness

New tests validate that the fast rectangle path and the C++ polygon clipper
match the expected behavior of the original R implementation. These tests are
located in `tests/testthat/` and run as part of the normal test suite.

## Future directions

After these changes, the main remaining hotspot for `geolines()` in local
profiling is `plot.xy`. Additional speedups may be possible through caching of
projected or clipped geometry in higher-level plotting workflows.
