---
title: "Rectangle Codes (d2r family)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rectangle Codes (d2r family)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

```{r}
library(geo)
```

## Overview

The `d2r` family converts geographic coordinates (decimal degrees) to rectangle
codes used in traditional and grid-based systems, and the `r2d` family converts
those codes back to rectangle center coordinates.

- `d2r` / `r2d`: Icelandic 30' latitude × 1° longitude rectangles.
- `d2sr` / `sr2d`: Sub-rectangles (15' × 30') within the Icelandic system.
- `d2mr` / `mr2d`: Minute-based grids (user-specified `dlat`/`dlon` in minutes).
- `d2dr` / `dr2d`: Degree-based grids (user-specified `dlat`/`dlon` in degrees).

A small numerical offset is applied internally to avoid ambiguous border cases.

## Icelandic rectangle codes (d2r / r2d)

```{r}
# Convert point to rectangle code and back to the rectangle center
code <- d2r(lat = 65.25, lon = -19.5)
center <- r2d(code)
code
center
```

```{r}
# Tally coastline points into rectangles
data(island)
rects <- d2r(island)
head(table(rects))
```

## Sub-rectangles (d2sr / sr2d)

```{r}
# Sub-rectangle code and back to center
sr_code <- d2sr(lat = 65.25, lon = -19.5)
sr_center <- sr2d(sr_code)
sr_code
sr_center
```

## Minute-based grid codes (d2mr / mr2d)

```{r}
# 5' latitude by 10' longitude grid
mr_code <- d2mr(lat = 65.25, lon = -19.5, dlat = 5, dlon = 10)
mr_center <- mr2d(mr_code, dlat = 5, dlon = 10)
mr_code
mr_center
```

## Degree-based grid codes (d2dr / dr2d)

```{r}
# 1° latitude by 2° longitude grid, starting at 50°N
# (Use startLat to align the coding origin)
dr_code <- d2dr(lat = 65.25, lon = -19.5, dlat = 1, dlon = 2, startLat = 50)
dr_center <- dr2d(dr_code, dlat = 1, dlon = 2, startLat = 50)
dr_code
dr_center
```

## Round-trip checks

```{r}
# Round-trips should land at the rectangle center for the chosen system
r2d(d2r(lat = 65 + 1/4, lon = -19 - 1/2))
```

```{r}
# Rectangle code -> center -> back to code
# (Center stays in the same rectangle)
d2r(r2d(519))
```
